---
title: "Make-MESA-TOPMed-Protein-Table-README"
author: "LekkiWood (LekkiWood@Gmail.com)"
date: today
format:
  pdf:
    toc: true
    toc-depth: 3
crossref:
  tbl-prefix: "Table"
  fig-prefix: "Figure"
execute:
  echo: false
  warning: false
  message: false
  # ensure chunks run from the project root (where _targets/ lives)
  dir: project
pdf-engine: latexmk
editor: visual
---

```{r}
#| label: load-packages


## Load targets
if (!requireNamespace("targets", quietly = TRUE)) {
  stop("The {targets} package is not installed/available in this render session.")
}

## Load knitr
if (!requireNamespace("knitr", quietly = TRUE)) {
  stop("The {knitr} package is not installed/available in this render session.")
}

## Load quarto
if (!requireNamespace("quarto", quietly = TRUE)) {
  stop("The {quarto} package is not installed/available in this render session.")
}
```

This README file describes the process for taking the raw data from the TOPMed X01 the protein assays and creating the following:

-   File 1:

-   File 2:

-   File 3:

-   File 4:

-   File 5:

## Important notes:

-   You should always check you have the most recent files! I use dynamic dates, so this can be done by checking that your files match the versions on Github:

    -   My profile can be found [here](https://github.com/LekkiWood).

    -   My repositories are listed [here](https://github.com/LekkiWood?tab=repositories).

    -   The repository for this project can be found [here](https://github.com/LekkiWood/Make-MESA-TOPMed-Proteomics-Olink3K).

    -   The most recent version of this PDF README file can be found

    -   The most recent version of the source code for this README file can be found

    -   The most recent source code for creating the output files can be found

\newpage

# File 1: Formatted Protein Assay Table

```{r}

#| label: load-filename-info

QC_info <- targets::tar_read(build_proteins_QC)
```

## Input file names

### Raw metabolite tables

-   A table of metabolite abundances from the amide assay: `{r} print(unlist(strsplit(QC_info$filenames$raw_proteins_file_name, split="/"))[8])`

### Info files

-   Sample information to link TOPMed IDs to unique MESA SHARe ID and exam combinations: `{r} print(unlist(strsplit(QC_info$filenames$raw_protein_info_file_name, split="/"))[8])`

### Protein keys

-   Keys to link Olink IDs to names compounds: `{r} print(unlist(strsplit(QC_info$filenames$ raw_protein_keys_file_name, split="/"))[8])`

### Bridging file

-   A file to bridge SHARe ids (sidno) with MESA IDs (idno) `{r} print(unlist(strsplit(QC_info$filenames$raw_bridgingfile_file_name, split="/"))[7])`

## Raw file info

-   The raw protein abundance file contained information on N=`{r} print(QC_info$raw_inputs$raw_protein_N_all)` protein assays (including those included for QC).

-   When removing assays for QC, the raw protein abundance file contained information on N=`{r} print(QC_info$raw_inputs$raw_protein_N_proteins[1,1])` proteins.
